Periodograms for Multi-band Astronomical Time Series

### 2015-01-21 writeup/paper.tex
rework paper and figures

The most popular parametric method is the Lomb-Scargle periodogram (discussed in detail in \S~\ref{sec:periodograms}.
The Lomb-Scargle periodogram is related to the $\chi^2$ for a least-square fit of a single sinusoid to data
and can treat non-uniformly sampled time series with heteroscedastic measurement uncertainties. 
The underlying model of the Lomb–Scargle periodogram is nonlinear in frequency and basis functions at different
frequencies are not orthogonal. As a result, the periodogram has many local maxima and thus in practice the global 
maximum of the periodogram is found by grid search \citep[for details see, e.g.][]{ICVG2014}.
A more general parameteric method based on the use of continuous-time autoregressive moving average (CARMA) model
was recently introduced by \citep{Kelly14}. CARMA models can also treat non-uniformly sampled time series with 
heteroscedastic measurement uncertainties, and can handle complex variability patterns.

@@

The proposed method is essentially a generalization of the Lomb-Scargle method to 
multi-band case. The light curves in  each band are modeled as arbitrary truncated Fourier series, 
with the period and phase shared across all bands. The key aspect is the use of Tikhonov regularization 
(discussed in detail in \S \ref{regularization}) which drives most of the variability into the so-called base 
model common to all bands, while fits for individual bands describe residuals relative to the base model 
and typically require lower-order Fourier series. This decrease in the required number of fit parameters is the
main reason for improved performance.

In \S 2, we provide a brief overview of periodograms, including the matrix-based formulation used in this work. 
XXX Finish this paragraph after the paper is nearly completed...

@@

Because astronomical observing cadences are rarely so uniform, many have looked at extending the ideas behind the periodogram to work with unevenly-sampled data. Most famously, \citet{Lomb76} and \citet{Scargle82} extended earlier work to define the {\it normalized periodogram}:
\begin{equation}
  \eqlabel{LombScargle}
  P_N(\omega) = \frac{1}{2\,V}\left[
    \frac{\left[\sum_k(y_k - \bar{y})\cos\omega(t_k - \tau)\right]^2}
    {\sum_k \cos^2\omega(t_k - \tau)}
    +
    \frac{\left[\sum_k(y_k - \bar{y})\sin\omega(t_k - \tau)\right]^2}
    {\sum_k \sin^2\omega(t_k - \tau)}
\right],
\end{equation}
where $\bar{y}$ is the mean and $V$ is the variance of the data $\{y_k\}$, and $\tau$ is the time-offset which makes $P_N(\omega)$ independent of a translation in $t$ \citep[see][for an in-depth discussion]{NumRec}. \citet{Lomb76} showed that this offset has a deeper effect: namely, it makes $P_N$ identical to the estimate of harmonic content given a least-squares fit to a single-component sinusoidal model,

@@

@@

\subsubsection{Matrix Formalism}
The expressions related to the stationary sinusoid model can be expressed more compactly by defining the following matrices:
\begin{equation}
X_\omega = \left[
\begin{array}{cc}
\sin(\omega t_1) & \cos(\omega t_1)\\
\sin(\omega t_2) & \cos(\omega t_2)\\
\vdots & \vdots \\
\sin(\omega t_N) & \cos(\omega t_N)\\
\end{array}
\right];~~
y = \left[
\begin{array}{c}
y_1 \\
y_2\\
\vdots \\
y_N\\
\end{array}
\right];~~
\Sigma = \left[
\begin{array}{cccc}
\sigma_1^2 & 0 &  \cdots & 0\\
0 & \sigma_2^2 &  \cdots & 0\\
\vdots & \vdots &  \ddots & \vdots\\
0 & 0 &  \cdots & \sigma_N^2
\end{array}
\right]
\end{equation}
With these definitions, the model in \eq{simplemodel} can be expressed as a simple linear product: $y(t|\omega,\theta) = X_\omega\theta$, and the model and reference $\chi^2$ can be written

\begin{eqnarray}
  \chi^2(\omega) &=& (y - X_\omega\theta)^T\Sigma^{-1}(y - X_\omega\theta)\\
  \chi^2_0 &=& y^T \Sigma^{-1} y
\end{eqnarray}
The expression for the normalized periodogram can be computed by finding the value of $\theta$ which minizes $\chi^2(\omega)$ via standard methods and plugging the result into \eq{chi2PN} to find
\begin{equation}
  \eqlabel{LombScargle2}
  P_N(\omega) = \frac{y^T\Sigma^{-1}X_\omega~[X_\omega^T\Sigma^{-1}X_\omega]^{-1}~X_\omega^T\Sigma^{-1}y}{y^T\Sigma^{-1}y}.
\end{equation}
We note that this expression is equivalent to \eq{LombScargle} in the homoscedastic case with $\Sigma \propto V I$.

@@

The upper-right panel shows the normalized periodogram for this source as a function of period. While the power does peak at the true period of 0.622 days, an aliasing effect is readily apparent near $P=0.38$. This additional peak is due to beat frequency between the true period $P$ and the observing cadence of $\sim 1$ day. This beat frequency is the first in a large sequence: for nightly observations, we'd expect to find excess power at periods That is, for nightly observations, we'd expect excess power to be found at frequencies $P_n = P / (1 + nP)$ days, for any positive or negative integer $n$. The strong alias in \fig{basic_example} corresponds to the $n=1$ beat period $P_n=0.383$. Though it is possible to carefully correct for such aliasing via the estimated window function \citep[e.g.][]{Roberts87}, we'll ignore this detail in the current work.

The lower-right panel of \fig{basic_example} shows the maximum likelihood interpretation of this periodogram: it is a measure of the normalized $\chi^2$ for a single-term sinusoidal model. Here we visualize the data from the left panel, but folded as a function of phase with the best-fit single-term model over-plot. Here its apparent that the single-term model is biased: RR Lyrae light curves are, in general, much more complicated than a simple sinusoid. Nevertheless, the simplistic sinusoidal model does well is able to recover the correct frequency to a high degree of accuracy (roughly related to the width of the peak) and significance (roughly related to the height of the peak). For a more complete introduction to and discussion of the single-term normalized periodogram, refer to, e.g. \citet{Bretthorst88} or \citet{ICVG2014}.

@@

The matrix-based formulation of \eq{LombScargle2}, though slower than the Fourier-derived formulation, is a more general expression and allows several advantages:
\begin{enumerate}
  \item It is trivially extended to heteroscedastic and/or correlated measurement noise in the data $y_k$ through appropriate modification of the noise covariance matrix $\Sigma$
  \item It is trivially extended to more sophisticated linear models by appropriately modifying the design matrix $X_\omega$.
  \item It is trivially extended to include Tikhonov/L2-regularization terms (see \S \ref{regularization} for more details)  by adding an appropriate diagonal term to the {\it normal matrix} $X_\omega^T\Sigma^{-1}X_\omega$.
\end{enumerate}
In the remainder of this section, we'll explore a few of these modifications and how they affect the periodogram and resulting model fits.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig02.pdf}
  \caption{
    An illustration of the effect of the floating mean model for biased data.
    The data consist of 80 observations drawn from a sinusoidal model. To mimick a potentially damaging selection effect, all observations with magnitude fainter than 16 are removed (indicated by the light-gray points). The standard and floating-mean periodograms are computed from the remaining data; these fits are shown over the data in the right panel. Because of this biased observing pattern, the mean of the observed data is not a good predictor of the true mean, and the standard model fails to recover the true period of 0.622 days, while the floating mean model still finds the correct period.
  }
  \figlabel{floating_mean}
\end{figure}

@@

For well-sampled data, there is usually very little difference between a standard periodogram and a floating-mean periodogram. Where this becomes important is if selection effects or observing cadences cause there to be preferentially more observations at certain phases of the light curve: a toy example demonstrating this situation is shown in \fig{floating_mean}. The data are drawn from a sinusoid with Gaussian errors, and data with a magnitude fainter than 16 are removed to simulate an observational bias (right panel). Because of this observational bias, the mean of the observed data are a poor predictor of the true mean, causing the standard fixed-mean method to poorly fit the data. The floating-mean approach is able to automatically adjust for this bias, resulting in a periodogram which readily detects the input period of 0.622 days (lower-right panel).

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig03.pdf}
  \caption{
    An illustration of a periodogram based on the truncated Fourier model.
    The data are the same as those in \fig{basic_example}. Note that the
    higher-order models will show periodogram peaks at multiples of the true
    fundamental frequency $P_0$: this is because for integer $n$ less than
    the number of Fourier terms in the model, $P_0$ is a higher harmonic of
    the model at $P=nP_0$.
  }
  \figlabel{multiterm_example}
\end{figure}

As mentioned above, the standard periodogram is equivalent to fitting a single-term stationary sinusoidal model to the data. A natural extension is to instead use a multiple-term sinusoidal model, with frequencies at integer multiples of the fundamental frequency. With $M$ Fourier terms, there are $2M + 1$ free parameters, and the model is given by
\begin{equation}
  y(t|\omega,\theta) = \theta_0 + \sum_{n=1}^M \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right].
\end{equation}
Because this remains a linear model, it can be easily accommodated into the matrix formalism above. For example, an $M = 2$-term floating-mean model can be constructed by building a design matrix $X_\omega$ with $2M + 1 = 5$ columns:
\begin{equation}
X_\omega^{(2)} = \left[
\begin{array}{ccccc}
1 & \sin(\omega t_1) & \cos(\omega t_1) & \sin(2\omega t_1) & \cos(2\omega t_1)\\
1 & \sin(\omega t_2) & \cos(\omega t_2) & \sin(2\omega t_2) & \cos(2\omega t_2)\\
1 & \sin(\omega t_3) & \cos(\omega t_3) & \sin(2\omega t_3) & \cos(2\omega t_3)\\
\vdots & \vdots & \vdots & \vdots & \vdots \\
1 & \sin(\omega t_N) & \cos(\omega t_N) & \sin(2\omega t_N) & \cos(2\omega t_N)\\
\end{array}
\right]
\end{equation}
Computing the power via \eq{LombScargle2} using $X_\omega^{(2)}$ will give the two-term periodogram. As $M$ grows, the size of the design matrix grows, but the periodogram can be computed in the same manner. \fig{multiterm_example} shows a few examples of this multiterm Fourier approach as applied to the simulated RR Lyrae light curve from \fig{basic_example}. There are several important features of this figure that give us insight into the subtleties of this type of multiterm fit.

First, we see in the right panel that all three models show a significant spike at the true period of $P_0 = 0.622$. The higher-order models, however, also show a a spike in power at $P_1 = 2 P_0$: the reason for this is that for and $M>1$-term model, the period $P_0$ is the first harmonic of a model with fundamental frequency $2P_0$, and the higher-order models contain the single-period result.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig04.pdf}
  \caption{
    The effect of regularization on a high-order model. The data is the same as
    those in \fig{basic_example}. We fit a 20-term truncated Fourier model to
    the data, with and without a regularization term. Without regularization,
    the model oscillates widely to fit the noise in the data. The
    regularization term effectively damps the higher-order Fourier modes and
    removes this oscillating behavior, leading to a model which is much more
    robust to noise in the data.
  }
  \figlabel{regularized_example}
\end{figure}

The previous sections raise the question: how complicated a model should we use? It is easily shown that as we add more terms to the fit, the model will come closer and closer to the observed data. When the number of model parameters equals the number of data points, we can expect to fit the data {\it perfectly} (though in most cases numerical inaccuracies in the fit prevent a truly perfect fit). For very high-order models, the result become very sensitive to the noise in the data, and we say we have {\it over-fit} the data. This can be addressed by explicitly truncating the series, but we can also use a {\it regularization} term to mathematically ensure a less complicated model.

A regularization term is an explicit penalty on the magnitude of the model parameters $\theta$, and can take a number of forms. For computational simplicity here we'll use an {\it L2 regularization} (also known as Tikhonov Regularization), which is a quadratic penalty term in the model parameters added to the $\chi^2$. Mathematically, this is equivalent to the Bayesian approach of using a Gaussian prior on the model parameters.

We encode our regularization in the matrix $\Lambda = {\rm diag}([\lambda_1, \lambda_2 \cdots \lambda_M])$ for a model with $M$ parameters, and our expression for ``regularized'' $\chi^2$ becomes
\begin{equation}
  \eqlabel{chi2reg}
  \chi_\Lambda^2(\omega) = (y - X_\omega\theta)^T\Sigma^{-1}(y - X_\omega\theta) + \theta^T\Lambda\theta
\end{equation}
Minimizing this regularized $\chi^2$, solving for $\theta$, and plugging into the expression for $P_N$ gives us the regularized counterpart of \eq{LombScargle2}:
\begin{equation}
  \eqlabel{LombScargleReg}
  P_N(\omega) = \frac{y^T\Sigma^{-1}X_\omega~[X_\omega^T\Sigma^{-1}X_\omega + \Lambda]^{-1}~X_\omega^T\Sigma^{-1}y}{y^T\Sigma^{-1}y}.
\end{equation}
Notice that the effect of this regularization term is to add a diagonal penalty to the normal matrix $X_\omega^T\Sigma^{-1}X_\omega$, which has the additional feature that it can correct ill-posed models where the normal matrix is not invertible. This feature of the regularization will become important below.

@@

\section{A Multiple-Band Model}
To compute a periodogram for multi-band data, we'll take advantage of many of the nice features of the matrix form of the normalized periodogram, which allows us to specify an arbitrary linear model covering our data. We will construct a multi-term Fourier model with the following features:
\begin{enumerate}
  \item An $M_{base}$-term truncated Fourier fit which models a latent parameter, which we'll call the ``overall variability''.
  \item A set of $M_{band}$-term truncated Fourier fits, each of which models the residual of a single band from this overall variability.
\end{enumerate}
The total number of parameters for $K$ filters is then $M_K = (2M_{base} + 1) + K(2M_{band} + 1)$. As a result, for each band $b$ we have the following model of the observed magnitudes:
\begin{eqnarray}
  y_b(t|\omega,\theta) = &\theta_0 + \sum_{n=1}^{M_{base}} \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right]& +\\ 
  &\theta^{(b)}_0 + \sum_{n=1}^{M_{band}} \left[\theta^{(b)}_{2n - 1}\sin(n\omega t) + \theta^{(b)}_{2n}\cos(n\omega t)\right].&
\end{eqnarray}
The important feature of this model is that {\it all bands} share the same base parameters $\theta$, while their offsets $\theta^{(b)}$ are determined individually.

We can construct the normalized periodogram for this model by building a sparse design matrix with $M_K$ columns. Each row corresponds to a single observation through a single band. Columns corresponding to the base model and the matching observation band will have nonzero entries; all other columns will be filled with zeros. For example, the $N_{base}=1$ and $N_{band}=0$ model corresponds to one with a simple single-term periodic base frequency, and an independent constant offset term in each band. The associated design matrix will look as follows:
\begin{equation}
X_\omega^{(1,0)} = \left[
\begin{array}{cccccccc}
1 & \sin(\omega t_1) & \cos(\omega t_1) & 1 & 0 & 0 & 0 & 0\\
1 & \sin(\omega t_2) & \cos(\omega t_2) & 0 & 1 & 0 & 0 & 0\\
1 & \sin(\omega t_3) & \cos(\omega t_3) & 0 & 0 & 0 & 1 & 0\\
1 & \sin(\omega t_4) & \cos(\omega t_4) & 0 & 0 & 1 & 0 & 0\\
\vdots & \vdots & \vdots & & & \vdots & &\\
1 & \sin(\omega t_N) & \cos(\omega t_N) & 0 & 1 & 0 & 0 & 0\\
\end{array}
\right]
\end{equation}
Here the nonzero entries of the final five columns indicate the $ugriz$ band of the given observation: the first row is a $u$-band measurement, the second is a $g$-band, the third is a $i$-band, etc.

@@

We'll choose the latter here, and regularize all the band columns while leaving the base columns un-regularized: for the above $X_\omega$ matrix, this regularization will look like
\begin{equation}
  \Lambda^{(1,0)} = {\rm diag}([0, 0, 0, \epsilon, \epsilon, \epsilon, \epsilon, \epsilon])
\end{equation}
where $\epsilon$ is some small fraction of the trace of the normal matrix $[X_\omega^T\Sigma^{-1}X_\omega]$. The logic of this choice of regularization is that any component of the model which is common to all bands will be accounted for in the base terms: only components of the fit that differ in each band will affect the band terms. Setting $\epsilon$ to some small fraction of the trace ensures that the regularization effect will be small. With this regularization in place, the model is well-posed and \eq{LombScargleReg} can be used to straightforwardly compute the power. We'll see an example of this in action in \sect{Simulated}.

@@

\subsection{Relationship of Multiband and Single-band approaches}
\sectlabel{relationship}
There is a special case of the multi-band approach when $N_{base}=0$ and $N_{band}=1$. In this case, the base offset can be removed completely and the $X$ matrix can be arranged to be block-diagonal. A block-diagonal design matrix in a linear model indicates that components of the model are being solved independently: here the independent components are a standard single-band generalized periodogram within each of the $K$ bands.

>>>>

The most popular parametric method is the Lomb-Scargle periodogram, which is discussed in detail in \S~\ref{sec:periodograms}.
The Lomb-Scargle periodogram is related to the $\chi^2$ for a least-square fit of a single sinusoid to data
and can treat non-uniformly sampled time series with heteroscedastic measurement uncertainties. 
The underlying model of the Lomb–Scargle periodogram is nonlinear in frequency and basis functions at different
frequencies are not orthogonal. As a result, the periodogram has many local maxima and thus in practice the global 
maximum of the periodogram is found by grid search \citep[for details see, e.g.][]{ICVG2014}.
A more general parameteric method based on the use of continuous-time autoregressive moving average (CARMA) model
was recently introduced by \citep{Kelly14}. CARMA models can also treat non-uniformly sampled time series with 
heteroscedastic measurement uncertainties, and can handle complex variability patterns.

@@

The proposed method is essentially a generalization of the Lomb-Scargle method to 
multi-band case. The light curves in  each band are modeled as arbitrary truncated Fourier series, 
with the period, and optionally the phase, shared across all bands. The key aspect enabling this approach is the use of Tikhonov regularization 
(discussed in detail in \S \ref{regularization}) which drives most of the variability into the so-called {\it base 
model} common to all bands, while fits for individual bands describe residuals relative to the base model 
and typically require lower-order Fourier series. This decrease in the required number of fit parameters is the
main reason for improved performance.

\todo{In \S 2, we provide a brief overview of periodograms, including the matrix-based formulation used in this work. In \S...}

@@

Because astronomical observing cadences are rarely so uniform, many have looked at extending the ideas behind the periodogram to work with unevenly-sampled data. Most famously, \citet{Lomb76} and \citet{Scargle82} extended earlier work to define the {\it normalized periodogram}:
\begin{equation}
  \eqlabel{LombScargle}
  P_N(\omega) = \frac{1}{2\,V}\left[
    \frac{\left[\sum_k(y_k - \bar{y})\cos\omega(t_k - \tau)\right]^2}
    {\sum_k \cos^2\omega(t_k - \tau)}
    +
    \frac{\left[\sum_k(y_k - \bar{y})\sin\omega(t_k - \tau)\right]^2}
    {\sum_k \sin^2\omega(t_k - \tau)}
\right],
\end{equation}
where $\bar{y}$ is the mean and $V$ is the variance of the data $\{y_k\}$, and $\tau$ is the time-offset which makes $P_N(\omega)$ independent of a translation in $t$ \citep[see][for an in-depth discussion]{NumRec}. \citet{Lomb76} showed that this time-offset has a deeper effect: namely, it makes $P_N$ identical to the estimate of harmonic content given a least-squares fit to a single-component sinusoidal model,

@@

While the important features of least squares frequency estimation via Lomb-Scargle periodograms has been discussed elsewhere, we will prevent a brief introduction to the subject in the following section.
The matrix-based formalism we develop here will make clear how the method can be extended to more sophisticated models, including regularized multi-band methods.

@@

\subsubsection{Matrix Formalism}
The expressions related to the stationary sinusoid model can be expressed more compactly by defining the following matrices:
\begin{equation}
X_\omega = \left[
\begin{array}{cc}
\sin(\omega t_1) & \cos(\omega t_1)\\
\sin(\omega t_2) & \cos(\omega t_2)\\
\vdots & \vdots \\
\sin(\omega t_N) & \cos(\omega t_N)\\
\end{array}
\right];~~
y = \left[
\begin{array}{c}
y_1 \\
y_2\\
\vdots \\
y_N\\
\end{array}
\right];~~
\Sigma = \left[
\begin{array}{cccc}
\sigma_1^2 & 0 &  \cdots & 0\\
0 & \sigma_2^2 &  \cdots & 0\\
\vdots & \vdots &  \ddots & \vdots\\
0 & 0 &  \cdots & \sigma_N^2
\end{array}
\right]
\end{equation}
With these definitions, the model in \eq{simplemodel} can be expressed as a simple linear product, $y(t|\omega,\theta) = X_\omega\theta$, and the model and reference $\chi^2$ can be written

\begin{eqnarray}
  \chi^2(\omega) &=& (y - X_\omega\theta)^T\Sigma^{-1}(y - X_\omega\theta)\\
  \chi^2_0 &=& y^T \Sigma^{-1} y
\end{eqnarray}
The expression for the normalized periodogram can be computed by finding via standard methods the value of $\theta$ which minizes $\chi^2(\omega)$, and plugging the result into \eq{chi2PN}. This yields
\begin{equation}
  \eqlabel{LombScargle2}
  P_N(\omega) = \frac{y^T\Sigma^{-1}X_\omega~[X_\omega^T\Sigma^{-1}X_\omega]^{-1}~X_\omega^T\Sigma^{-1}y}{y^T\Sigma^{-1}y}.
\end{equation}
We note that this expression is equivalent to \eq{LombScargle} in the homoscedastic case with $\Sigma \propto V I$.

@@

The upper-right panel shows the normalized periodogram for this source as a function of period. While the power does peak at the true period of 0.622 days, an aliasing effect is readily apparent near $P=0.38$. This additional peak is due to beat frequency between the true period $P$ and the observing cadence of $\sim 1$ day. This beat frequency is the first in a large sequence: for nightly observations, we'd expect to find excess power at periods $P_n = P / (1 + nP)$ days, for any integer $n$. The strong alias in \fig{basic_example} corresponds to the $n=1$ beat period $P_n=0.383$. Though it is possible to carefully correct for such aliasing by iteratively removing contributions from the estimated window function \citep[e.g.][]{Roberts87}, we'll ignore this detail in the current work.

The lower-right panel of \fig{basic_example} shows the maximum likelihood interpretation of this periodogram: it is a measure of the normalized $\chi^2$ for a single-term sinusoidal model. Here we visualize the data from the left panel, but folded as a function of phase with the best-fit single-term model over-plot. Here it is apparent that the single-term model is highly biased: RR Lyrae light curves are, in general, much more complicated than a simple sinusoid. Nevertheless, the simplistic sinusoidal model is able to recover the correct frequency to a high degree of accuracy (roughly related to the width of the peak) and significance (roughly related to the height of the peak). For a more complete introduction to and discussion of the single-term normalized periodogram, refer to, e.g. \citet{Bretthorst88} or \citet{ICVG2014}.

@@

The matrix-based formulation of \eq{LombScargle2}, though slower than the Fourier-derived formulation, is a more general expression and allows several advantages:
\begin{enumerate}
  \item It is trivially extended to heteroscedastic and/or correlated measurement noise in the data $y_k$ through appropriate modification of the {\it noise covariance matrix} $\Sigma$
  \item It is trivially extended to more sophisticated linear models by appropriately modifying the {\it design matrix} $X_\omega$.
  \item It is trivially extended to include Tikhonov/L2-regularization terms (see \S \ref{regularization} for more details)  by adding an appropriate diagonal term to the {\it normal matrix} $X_\omega^T\Sigma^{-1}X_\omega$.
\end{enumerate}
In the remainder of this section, we'll explore a few of these modifications and how they affect the periodogram and resulting model fits.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig02.pdf}
  \caption{
    An illustration of the effect of the floating mean model for biased data.
    The data consist of 80 observations drawn from a sinusoidal model. To mimick a potentially damaging selection effect, all observations with magnitude fainter than 16 are removed (indicated by the light-gray points). The standard and floating-mean periodograms are computed from the remaining data; these fits are shown over the data in the left panel. Because of this biased observing pattern, the mean of the observed data is not a good predictor of the true mean, and the standard fixed-mean model fails to recover the true period of 0.622 days, while the floating mean model still finds the correct period.
  }
  \figlabel{floating_mean}
\end{figure}

@@

For well-sampled data, there is usually very little difference between a standard periodogram and a floating-mean periodogram. Where this becomes important is if selection effects or observing cadences cause there to be preferentially more observations at certain phases of the light curve: a toy example demonstrating this situation is shown in \fig{floating_mean}. The data are drawn from a sinusoid with Gaussian errors, and data with a magnitude fainter than 16 are removed to simulate an observational bias (left panel). Because of this observational bias, the mean of the observed data are a poor predictor of the true mean, causing the standard fixed-mean method to poorly fit the data (upper-right panel). The floating-mean approach is able to automatically adjust for this bias, resulting in a periodogram which readily detects the input period of 0.622 days (lower-right panel).

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig03.pdf}
  \caption{
    An illustration of a periodogram based on several truncated Fourier models.
    The data are the same as those in \fig{basic_example}. Note that the
    higher-order models will generally show periodogram peaks at multiples
    of the true fundamental frequency $P_0$: this is because for integer $n$
    less than the number of Fourier terms in the model, $P_0$ is a higher
    harmonic of the model at $P=nP_0$. Additionally, the increased degrees of
    freedom in the higher-order models let them fit better at any frequency,
    which drives up the ``background'' level in the periodogram.
  }
  \figlabel{multiterm_example}
\end{figure}

As mentioned above, the standard periodogram is equivalent to fitting a single-term stationary sinusoidal model to the data. A natural extension is to instead use a multiple-term sinusoidal model, with frequencies at integer multiples of the fundamental frequency. With $M$ Fourier terms, there are $2M + 1$ free parameters, and the model is given by
\begin{equation}
  y(t|\omega,\theta) = \theta_0 + \sum_{n=1}^M \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right].
\end{equation}
Because this model remains linear in the parameters $\theta$, it can be easily accommodated into the matrix formalism above. For example, an $M = 2$-term floating-mean model can be constructed by building a design matrix $X_\omega$ with $2M + 1 = 5$ columns:
\begin{equation}
X_\omega^{(2)} = \left[
\begin{array}{ccccc}
1 & \sin(\omega t_1) & \cos(\omega t_1) & \sin(2\omega t_1) & \cos(2\omega t_1)\\
1 & \sin(\omega t_2) & \cos(\omega t_2) & \sin(2\omega t_2) & \cos(2\omega t_2)\\
1 & \sin(\omega t_3) & \cos(\omega t_3) & \sin(2\omega t_3) & \cos(2\omega t_3)\\
\vdots & \vdots & \vdots & \vdots & \vdots \\
1 & \sin(\omega t_N) & \cos(\omega t_N) & \sin(2\omega t_N) & \cos(2\omega t_N)\\
\end{array}
\right]
\end{equation}
Computing the power via \eq{LombScargle2} using $X_\omega^{(2)}$ will give the two-term periodogram. As $M$ grows, the size of the design matrix grows, but the periodogram can be computed in the same manner. \fig{multiterm_example} shows a few examples of this multiterm Fourier approach as applied to the simulated RR Lyrae light curve from \fig{basic_example}. There are several important features of this figure that give us insight into the subtleties of this type of multiterm fit.

First, we see in the right panel that all three models show a clear spike at the true period of $P_0 = 0.622$. The higher-order models, however, also show a a spike in power at $P_1 = 2 P_0$: the reason for this is that for and $M>1$-term model, the period $P_0$ is the first harmonic of a model with fundamental frequency $2P_0$, and the higher-order models contain the single-period result.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig04.pdf}
  \caption{
    The effect of regularization on a high-order model. The data is the same as
    those in \fig{basic_example}. We fit a 20-term truncated Fourier model to
    the data, with and without a regularization term. Without regularization,
    the model oscillates widely to fit the noise in the data. The
    regularization term effectively damps the higher-order Fourier modes and
    removes this oscillating behavior, leading to a more robust model with
    stronger periodogram peaks.
  }
  \figlabel{regularized_example}
\end{figure}

The previous sections raise the question: how complicated a model should we use? We have seen that as we add more terms to the fit, the model will come closer and closer to the observed data. When the number of model parameters equals the number of data points, we can expect to fit the data {\it perfectly} (though in most cases numerical inaccuracies in the fit prevent a truly perfect fit). For very high-order models, such a fit becomes very sensitive to the noise in the data, and we say we have {\it over-fit} the data. This can be addressed by explicitly truncating the series, but we can also use a {\it regularization} term to mathematically ensure a less complicated model.

A regularization term is an explicit penalty on the magnitude of the model parameters $\theta$, and can take a number of forms. For computational simplicity here we'll use an {\it L2 regularization} -- also known as Tikhonov Regularization \citep{Tikhonov1963} or Ridge Regression \citep{Hoerl1970} -- which is a quadratic penalty term in the model parameters added to the $\chi^2$. Mathematically, this is equivalent to the Bayesian approach of using a zero-mean Gaussian prior on the model parameters.

We encode our regularization in the matrix $\Lambda = {\rm diag}([\lambda_1, \lambda_2 \cdots \lambda_M])$ for a model with $M$ parameters, and our expression for ``regularized'' $\chi^2$ becomes
\begin{equation}
  \eqlabel{chi2reg}
  \chi_\Lambda^2(\omega) = (y - X_\omega\theta)^T\Sigma^{-1}(y - X_\omega\theta) + \theta^T\Lambda\theta
\end{equation}
Minimizing this regularized $\chi^2$, solving for $\theta$, and plugging into the expression for $P_N$ gives us the regularized counterpart of \eq{LombScargle2}:
\begin{equation}
  \eqlabel{LombScargleReg}
  P_N(\omega) = \frac{y^T\Sigma^{-1}X_\omega~[X_\omega^T\Sigma^{-1}X_\omega + \Lambda]^{-1}~X_\omega^T\Sigma^{-1}y}{y^T\Sigma^{-1}y}.
\end{equation}
Notice that the effect of this regularization term is to add a diagonal penalty to the normal matrix $X_\omega^T\Sigma^{-1}X_\omega$, which has the additional feature that it can correct ill-posed models where the normal matrix is non-invertible. This feature of the regularization will become important for the multiband models discussed below.

@@

\section{A Multiple-Band Model}
\sectlabel{multiband}
To compute a periodogram for multi-band data, we'll take advantage of many of the nice features of the matrix form of the normalized periodogram, which easily accommodates arbitrary linear models covering our data.
To account for data in multible bands, we will construct a multi-term Fourier model with the following features:
\begin{enumerate}
  \item An $M_{base}$-term truncated Fourier fit which models a latent parameter, which we'll call the ``overall variability''.
  \item A set of $M_{band}$-term truncated Fourier fits, each of which models the residual of a single band from this overall variability.
\end{enumerate}
The total number of parameters for $K$ filters is then $M_K = (2M_{base} + 1) + K(2M_{band} + 1)$. As a result, for each band $b$ we have the following model of the observed magnitudes:
\begin{eqnarray}
  y_b(t|\omega,\theta) = &\theta_0 + \sum_{n=1}^{M_{base}} \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right]& +\\ 
  &\theta^{(b)}_0 + \sum_{n=1}^{M_{band}} \left[\theta^{(b)}_{2n - 1}\sin(n\omega t) + \theta^{(b)}_{2n}\cos(n\omega t)\right].&
\end{eqnarray}
The important feature of this model is that {\it all bands} share the same base parameters $\theta$, while their offsets $\theta^{(b)}$ are determined individually.

We can construct the normalized periodogram for this model by building a sparse design matrix with $M_K$ columns. Each row corresponds to a single observation through a single band. Columns corresponding to the base model and the matching observation band will have nonzero entries; all other columns will be filled with zeros. For example, the $N_{base}=1$ and $N_{band}=0$ model corresponds to one with a simple single-term periodic base frequency, and an independent constant offset term in each band. The associated design matrix will look as follows:
\begin{equation}
X_\omega^{(1,0)} = \left[
\begin{array}{cccccccc}
1 & \sin(\omega t_1) & \cos(\omega t_1) & 1 & 0 & 0 & 0 & 0\\
1 & \sin(\omega t_2) & \cos(\omega t_2) & 0 & 1 & 0 & 0 & 0\\
1 & \sin(\omega t_3) & \cos(\omega t_3) & 0 & 0 & 0 & 1 & 0\\
\vdots & \vdots & \vdots & & & \vdots & &\\
1 & \sin(\omega t_N) & \cos(\omega t_N) & 0 & 0 & 1 & 0 & 0\\
\end{array}
\right]
\end{equation}
Here the nonzero entries of the final five columns are binary flags indicating the $ugriz$ band of the given observation: in this case, the first row is a $u$-band measurement, the second is a $g$-band, the third is a $i$-band, etc.

@@

We'll choose the latter here, and regularize all the band columns while leaving the base columns un-regularized: for the above $X_\omega$ matrix, this regularization will look like
\begin{equation}
  \Lambda^{(1,0)} = {\rm diag}([0, 0, 0, \epsilon, \epsilon, \epsilon, \epsilon, \epsilon])
\end{equation}
where $\epsilon$ is some small fraction of the trace of the normal matrix $[X_\omega^T\Sigma^{-1}X_\omega]$. The logic of this choice of regularization is that any component of the model which is common to all bands will be preferentially reflected in the base terms, with independent behavior reflected in the individual band terms. Setting $\epsilon$ to some small fraction of the trace ensures that the regularization effect on the remaining model will be small. With this regularization in place, the model is well-posed and \eq{LombScargleReg} can be used to straightforwardly compute the power. We'll see an example of this in action in \sect{Simulated}.

@@

This multiband approach, then, actually comprises a set of models indexed by their value of $N_{base}$ and $N_{band}$. The most fundamental models have $(N_{base}, N_{band}) = (1,0)$ and $(0,1)$, which we'll call the {\it shared-phase} and {\it multi-phase} models respectively. In the shared-phase model, all variability is assumed to be shared between the bands, with only the fixed offset between them allowed to float. In the multi-phase model, each band has independent variability around a shared fixed offset.

\subsection{Relationship of Multiband and Single-band approaches}
\sectlabel{relationship}
The multi-phase $N_{base}=0$, $N_{band}=1$ model turns out to be a particularly special case.
In it, the base offset can be removed completely and the design matrix $X_\omega$ can be arranged to be block-diagonal. A block-diagonal design matrix in a linear model indicates that components of the model are being solved independently: here these independent components are a standard single-band generalized periodogram fit within each of the $K$ bands.
